---
title: "GPTAnno Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPTAnno Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette walks through the main `GPTAnno` workflow at a high level:

1. prepare a Seurat object
2. build Cell Ontology support objects
3. run multi-resolution clustering and marker detection
4. annotate clusters across resolutions with repeated LLM calls
5. score resolutions and select the best parent annotation
6. optionally subcluster and annotate finer subtypes

This is an **illustrated pipeline vignette**, not a heavy end-to-end analysis
with bundled real data. That keeps the vignette lightweight, reproducible, and
safe to build without requiring:

- large example datasets
- live LLM API calls during vignette build
- long clustering runtimes

For helper utilities such as ontology search, CL relationship inspection, and
agreement scoring, see the separate vignette:

- `vignette("helper-utilities", package = "GPTAnno")`

## Why an Illustrated Pipeline Instead of Real Data?

For this package, an illustrated pipeline is the better default vignette:

- the main workflow depends on user data, API keys, and live LLM responses
- a real dataset vignette would either be slow and expensive or too fragile for
  package builds
- most users first need the workflow structure and the required objects/inputs

A real-data analysis is still valuable, but it fits better as:

- a separate long-form tutorial outside the package build
- a README-linked notebook
- a website article or reproducible analysis repo

## Setup

```{r}
library(GPTAnno)
library(Seurat)
library(ontologyIndex)
```

Choose an LLM backend by setting the corresponding API key or by using a local
provider through `llm_config`.

```{r eval = FALSE}
Sys.setenv(OPENAI_API_KEY = "your-api-key")

llm_config <- list(
  provider = "openai",
  model = "gpt-5",
  params = ellmer::params(temperature = 0)
)
```

Other supported providers use the same structure:

```{r eval = FALSE}
llm_config <- list(provider = "anthropic", model = "claude-sonnet-4-20250514")
llm_config <- list(provider = "gemini", model = "gemini-2.0-flash")
```

## Step 1: Build Ontology Objects

Many GPTAnno features depend on the Cell Ontology graph.

```{r eval = FALSE}
cl <- get_ontology("http://purl.obolibrary.org/obo/cl.obo", extract_tags = "everything")
graph <- build_ontology_graph(cl)
```

These objects are used for:

- ontology-aware prompting
- distance calculations
- CL relationship inspection

## Step 2: Prepare the Seurat Object

If your Seurat object is already normalized and reduced, you can move directly
to clustering. Otherwise:

```{r eval = FALSE}
seurat_obj <- preprocess_seurat_object(
  seurat_obj,
  npcs = 30
)
```

## Step 3: Run Multi-Resolution Clustering and Marker Detection

GPTAnno expects marker files for each tested resolution.

```{r eval = FALSE}
marker_dir <- "output/marker_genes"

cluster_results <- run_multi_resolution_clustering(
  seurat_obj = seurat_obj,
  resolutions = c(0.1, 0.5, 0.9, 1.2),
  result_dir = marker_dir,
  dims = 1:30
)

seurat_obj <- cluster_results$seurat_obj
```

This step writes marker files such as:

- `markers_res_0.1.rds`
- `markers_res_0.5.rds`

Based on your dataset, we recommend starting with a relatively wide range of
resolutions, for example `c(0.1, 0.5, 0.9, 1)`, and then refining if needed.

## Step 4: Parent-Level Annotation Across Resolutions

Run repeated LLM annotation for each candidate clustering resolution.

```{r eval = FALSE}
annotation_results <- gptanno(
  seurat_obj = seurat_obj,
  resolutions = c(0.1, 0.5, 0.9, 1.2),
  cl = cl,
  graph = graph,
  tissue_name = "mouse adult heart",
  llm_config = llm_config,
  marker_dir = marker_dir,
  n_runs = 10,
  topgenenumber = 10
)

saveRDS(annotation_results, "annotation_results.rds")
```

Each resolution returns:

- raw repeated predictions across runs
- per-cluster summarized annotations
- final selected annotation per cluster
- ontology distance summaries when available

We recommend saving `annotation_results` for future reference, because it keeps
the raw repeated predictions across runs. If needed, users should also inspect
the marker files in `marker_dir`, since those are the direct inputs used for the
LLM prompts.

## Step 5: Score Resolutions and Pick the Best Parent Annotation

Use the package composite score to compare resolution-level annotation quality.

```{r eval = FALSE}
scores <- score_annotation_resolutions(annotation_results)
scores

# Optional: save the score table as a CSV
scores <- score_annotation_resolutions(
  annotation_results,
  output_csv = "annotation_composite_score.csv"
)
```

Providing `output_csv` is optional, but useful if you want a saved score table
for downstream review.

The best resolution is the first row after sorting by `composite_score`.

```{r eval = FALSE}
best_res_name <- scores$resolution[1]
best_res_obj <- annotation_results[[best_res_name]]
res_val <- sub("^res_", "", best_res_name)
cluster_col <- paste0("cluster_res.", res_val)
```

Assign the chosen parent annotation back to the Seurat object:

```{r eval = FALSE}
seurat_obj <- assign_celltype(
  seurat_obj = seurat_obj,
  annotation_summary = best_res_obj,
  cluster_col = cluster_col,
  new_celltype = "celltype_parent"
)
```

## Step 6: Optional Subclustering

After assigning broad parent cell types, you can refine selected populations.

There are two ways to choose which parent cell types will be subclustered:

- **Automatic selection by GPTAnno**: the package selects parent cell types
  based on built-in criteria such as cell count and clustering structure
- **User-specified selection**: the user explicitly passes
  `celltypes_to_subcluster`

### Option A: Let GPTAnno Select Parent Cell Types to Subcluster

```{r eval = FALSE}
subcluster_results <- subcluster_and_find_markers(
  seurat_obj = seurat_obj,
  cl = cl,
  predicted_celltype_column = "celltype_parent",
  output_dir = "output/subclusters",
  resolutions = c(0.1, 0.2, 0.3),
  dims = 1:30
)
```

### Option B: User-Specified Parent Cell Types to Subcluster

```{r eval = FALSE}
subcluster_results <- subcluster_and_find_markers(
  seurat_obj = seurat_obj,
  cl = cl,
  predicted_celltype_column = "celltype_parent",
  output_dir = "output/subclusters",
  resolutions = c(0.1, 0.2, 0.3),
  dims = 1:30,
  celltypes_to_subcluster = c("T cell", "B cell", "endothelial cell")
)
```

Use `celltypes_to_subcluster` when you already know which parent populations
you want to refine, or when you want to override the package's automatic
selection logic.

## Step 7: Choose a Subcluster Annotation Strategy

GPTAnno supports two complementary subcluster strategies.

### Ontology-Constrained Strategy

Use this when you want predictions restricted to Cell Ontology descendants.

```{r eval = FALSE}
subcluster_workflow_results <- run_subcluster_annotation_workflow(
  base_dir = "output/subclusters",
  strategy = "ontology",
  cl = cl,
  tissue_name = "mouse adult heart",
  resolutions = c(0.1, 0.2, 0.3),
  llm_config = llm_config,
  n_runs = 10,
  select_best = TRUE
)
```

### Marker-Inheritance Strategy

Use this when you want more discovery-oriented subtype annotation.

```{r eval = FALSE}
subcluster_workflow_results <- run_subcluster_annotation_workflow(
  base_dir = "output/subclusters",
  strategy = "marker_inheritance",
  cl = cl,
  parent_marker_root = marker_dir,
  parent_res = res_val,
  parent_cluster_col = cluster_col,
  tissue_name = "mouse adult heart",
  llm_config = llm_config,
  n_runs = 10
)
```

## Step 8: Assign Subcluster Results Back to the Full Object

You can either use the best subcluster resolution chosen by the workflow, or
manually decide which resolution to use for each parent cell type.

### Option A: Use the Best Resolution Chosen by GPTAnno

```{r eval = FALSE}
seurat_obj <- assign_subcluster_annotations_to_full(
  full_seurat = seurat_obj,
  workflow_results = subcluster_workflow_results,
  use_best = TRUE,
  parent_column = "celltype_parent"
)
```

### Option B: Manually Choose the Resolution per Parent Cell Type

```{r eval = FALSE}
seurat_obj <- assign_subcluster_annotations_to_full(
  full_seurat = seurat_obj,
  workflow_results = subcluster_workflow_results,
  use_best = FALSE,
  resolution_map = c(
    "T cell" = "res_0.2",
    "B cell" = "res_0.1"
  ),
  parent_column = "celltype_parent"
)
```

## Evaluating Predictions Against Manual Labels

If you have a reference annotation column, GPTAnno provides ontology-aware
evaluation functions.

```{r eval = FALSE}
ancestor_type_map <- build_ancestor_type_map(cl)

agreement <- score_annotation_agreement_ontology(
  seurat_obj,
  manual_col = "celltype_manual",
  predicted_col = "celltype_parent",
  cl_term_map = GPTAnno::cl_term_map,
  ancestor_type_map = ancestor_type_map
)

agreement$summary
```

You can also inspect ontology distance:

```{r eval = FALSE}
distance_res <- score_annotation_distance_ontology(
  seurat_obj,
  manual_col = "celltype_manual",
  predicted_col = "celltype_parent",
  graph = graph
)
```

## Practical Notes

- Use the same `marker_dir` for clustering output and `gptanno()` input.
- If some clusters have no usable marker genes after filtering, GPTAnno now
  leaves them as `"unknown"` and still keeps valid predictions for the other
  clusters in the batch.
- `llm_config$params` is passed directly to `ellmer`, so model-specific decoding
  settings should be checked against the provider documentation.
- For local models, use `check_ollama_available()` and `list_ollama_models()`
  before launching a pipeline run.

## Recommended Documentation Split

For this package, the most maintainable documentation split is:

- **README**: short landing page and quick start
- **Pipeline vignette**: main package workflow, illustrated but not heavy
- **Helper utilities vignette**: ontology/search/scoring utilities outside the
  main pipeline

That gives users both a clear entry point and a reusable reference layer
without forcing live LLM calls or a large bundled dataset into the package
build.
