---
title: "Helper Utilities in GPTAnno"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Helper Utilities in GPTAnno}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`GPTAnno` includes a set of helper functions that are useful even outside the
main parent/subcluster annotation workflow. They fall into a few practical
groups:

- annotation cleaning and ontology lookup
- ontology graph and ancestry helpers
- relationship and distance inspection
- agreement and resolution scoring
- LLM provider utilities

This vignette gives a compact tour of the helpers that users are most likely to
call directly.

## Annotation Cleaning and Lookup

`clean_and_match_annotation()` standardizes raw labels and tries to map them to
known Cell Ontology naming.

```{r}
library(GPTAnno)

clean_and_match_annotation("Helper T cells")
clean_and_match_annotation("cardiomyocytes")
clean_and_match_annotation("unknown cell")
```

`map_celltypes_to_cl()` maps cleaned labels to CL IDs and official labels using
the built-in term map.

```{r}
map_celltypes_to_cl(c("fibroblast", "ciliated epithelial cell"), verbose = FALSE)
```

`search_ols()` queries the EMBL-EBI Ontology Lookup Service for Cell Ontology
terms. This is useful when an exact local match is unavailable or when you want
suggestions.

```{r eval = FALSE}
search_ols("fibroblast")
search_ols("lymphatic endothelial cell", size = 3)
search_ols("ciliated epithelial", size = 3)
```

## Ontology Setup

Several helper functions need a Cell Ontology object plus graph-based metadata.

```{r eval = FALSE}
library(ontologyIndex)

cl <- get_ontology("http://purl.obolibrary.org/obo/cl.obo", extract_tags = "everything")
graph <- build_ontology_graph(cl)
ancestor_type_map <- build_ancestor_type_map(cl)
cl_map <- build_cl_term_map(cl)
```

`build_ontology_graph()` converts the ontology parent structure into an
`igraph`, and `build_ancestor_type_map()` stores all ancestors for each CL ID.

## Inspecting Relationships Between Cell Types

`check_cl_relationship()` compares two CL IDs and reports whether they are
exact, parent/child, sibling, or unrelated.

```{r eval = FALSE}
check_cl_relationship("CL:0000066", "CL:0000312", cl, graph)
```

`get_common_ancestors()` is useful when comparing two labels and asking what
higher-level ontology terms they share.

```{r eval = FALSE}
get_common_ancestors(
  "helper T cell",
  "cytotoxic T cell",
  cl_term_map = cl_map,
  cl_ontology = cl,
  ancestor_type_map = ancestor_type_map
)
```

The package also exposes lower-level ontology inspection helpers:

```{r eval = FALSE}
get_cl_direct_parents("CL:0000067", cl)
get_cl_direct_children("CL:0000115", cl)
get_cl_node_depths(c("CL:0000067", "CL:0000115"), graph)
get_cl_pairwise_distances(
  data.frame(from = "CL:0000067", to = "CL:0000115"),
  graph
)
get_cl_names(c("CL:0000067", "CL:0000115"), cl)
```

## Scoring Annotation Results

`score_annotation_resolutions()` compares multiple annotation summaries and
computes the package composite score used to rank resolutions or strategies.

```{r eval = FALSE}
score_annotation_resolutions(annotation_result_list)
```

For manual-versus-predicted comparison on a Seurat object:

- `score_annotation_distance_ontology()` measures ontology path distance.
- `score_annotation_agreement_ontology()` uses ancestor-aware agreement scoring.
- `score_annotation_agreement_ontology_detailed()` returns more detailed match
  categories.

```{r eval = FALSE}
score_annotation_distance_ontology(
  seurat_obj,
  manual_col = "manual_celltype",
  predicted_col = "predicted_celltype",
  graph = graph
)

score_annotation_agreement_ontology(
  seurat_obj,
  manual_col = "manual_celltype",
  predicted_col = "predicted_celltype",
  ancestor_type_map = ancestor_type_map
)
```

## LLM Utilities

If you want to test GPTAnno's provider layer directly, the package exposes:

- `call_llm()` for direct provider calls
- `prepare_config()` for merging legacy `model` and `llm_config`
- `check_ollama_available()` and `list_ollama_models()` for local Ollama use

```{r eval = FALSE}
cfg <- prepare_config(
  model = "gpt-5",
  llm_config = list(provider = "openai", model = "gpt-5")
)

call_llm(
  prompt = "Reply with the word fibroblast.",
  provider = cfg$provider,
  model = cfg$model
)
```

## Visualization

`plot_celltype_comparison()` gives a quick side-by-side visualization of
original clusters and assigned annotations in a Seurat object.

```{r eval = FALSE}
plot_celltype_comparison(
  seurat_obj,
  original_col = "seurat_clusters",
  annotation_col = "annotated_celltype"
)
```

## Notes

- `search_ols()` depends on the live OLS API, so it may return different
  results over time.
- Functions that use CL IDs or ontology relationships work best when you load a
  current `cl.obo` file and build `graph` and `ancestor_type_map` from it.
- The built-in `cl_term_map` is convenient for day-to-day use, but
  `build_cl_term_map()` is available if you want a fresh map from a local
  ontology snapshot.
